---
title: "2020 NZ Election"
output: 
  flexdashboard::flex_dashboard:
    theme: spacelab
    self_contained: false
    css: https://fonts.googleapis.com/css?family=Open%20Sans
    lib_dir: lib
    mathjax: null
    navbar:
      - { title: "Intro", icon: "fa-home", href: "index.html", align: left }
      - { title: "Parties", icon: "ion-ios-settings-strong", href: "dumbbells.html", align: left }
      - { title: "Refs", icon: "ion-ios-settings-strong", align: left }
      - { title: "Hexmaps", icon: "ion-cube", href: "hexmaps.html", align: left }
      - { title: "@dakvid", icon: "fa-twitter", href: "https://twitter.com/dakvid", align: right }
    social: menu
    source_code: https://github.com/dakvid/election2020
editor_options: 
  chunk_output_type: console
---
  
```{r init, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r setup, include=FALSE}
library(magrittr)
library(readxl)
library(dplyr)
library(forcats)
library(glue)
library(stringr)
library(plotly)

ordinary <- 
  read_xlsx("data/ElectorateReferendumResults.xlsx",
            sheet = "Ordinary Votes",
            skip = 6) %>% 
  set_names(c("id", "gm_name",
              "euth_yes_ordinary", "euth_no_ordinary", "euth_inf_ordinary",
              "canna_yes_ordinary", "canna_no_ordinary", "canna_inf_ordinary",
              "total_ordinary")) %>% 
  mutate(id = id %>% str_extract("^[^/]+") %>% as.numeric())

special <- 
  read_xlsx("data/ElectorateReferendumResults.xlsx",
            sheet = "Special Votes",
            skip = 5) %>% 
  set_names(c("id", "electorate_name",
              "euth_yes_special", "euth_no_special", "euth_inf_special",
              "canna_yes_special", "canna_no_special", "canna_inf_special",
              "total_special")) %>% 
  mutate(id = as.numeric(id))

total_votes <- 
  add(
    ordinary %>% 
      pull(total_ordinary) %>% 
      sum(),
    special %>% 
      pull(total_special) %>% 
      sum()
  )

maori_specials_total <- 
  special %>% 
  anti_join(ordinary,
            by = "id") %>% 
  pull(total_special) %>% 
  sum()

maori_specials <- 
  special %>% 
  anti_join(ordinary,
            by = "id") %>% 
  mutate(id = 99,
         electorate_name = "M훮ori special votes") %>% 
  group_by(id, electorate_name) %>% 
  summarise(euthenasia = sum(euth_yes_special),
            cannabis = sum(canna_yes_special),
            total = sum(total_special)) %>% 
  ungroup() %>% 
  transmute(id,
            electorate_name,
            euthenasia = round(100 * euthenasia / total, 1),
            cannabis = round(100 * cannabis / total, 1),
            difference = abs(euthenasia - cannabis))

  electorate_votes <- 
    ordinary %>% 
    inner_join(special,
               by = "id") %>% 
    transmute(id,
              electorate_name,
              euthenasia = round(100 * (euth_yes_ordinary + euth_yes_special) / (total_ordinary + total_special), 1),
              cannabis = round(100 * (canna_yes_ordinary + canna_yes_special) / (total_ordinary + total_special), 1),
              difference = abs(euthenasia - cannabis))

  referenda <- 
    bind_rows(electorate_votes,
              maori_specials) %>% 
    mutate(electorate_name = if_else(is.na(id), "New Zealand", electorate_name),
           id = if_else(is.na(id), 0, id),
           c_compare = if_else(cannabis > euthenasia, "more", "fewer"),
           e_compare = if_else(euthenasia > cannabis, "more", "fewer"))
```


# 


## Plots {.tabset}

### Cannabis

```{r db_cannabis}
referenda %>% 
  mutate(electorate_name = fct_reorder(electorate_name, cannabis)) %>% 
  plot_ly(color = I("gray80")) %>% 
  add_markers(
    x = ~cannabis, y = ~electorate_name, name = "cannabis",
    color = I("green"),
    hoverinfo = "text",
    text = ~glue("{cannabis}% of {electorate_name} voters supported")
  ) %>% 
  layout(
    title = glue("Cannabis Referendum Support by Electorate"),
    xaxis = list(title = "Votes (Percent)"),
    yaxis = list(title = "", dtick = 1),
    margin = list(l = 130)
  )
```

### Euthenasia

```{r db_euthenasia}
referenda %>% 
  mutate(electorate_name = fct_reorder(electorate_name, euthenasia)) %>% 
  plot_ly(color = I("gray80")) %>% 
  add_markers(
    x = ~euthenasia, y = ~electorate_name, name = "euthenasia",
    color = I("gray10"),
    hoverinfo = "text",
    text = ~glue("{euthenasia}% of {electorate_name} voters supported")
  ) %>% 
  layout(
    title = glue("Euthenasia Referendum Support by Electorate"),
    xaxis = list(title = "Votes (Percent)"),
    yaxis = list(title = "", dtick = 1),
    margin = list(l = 130)
  )
```

### Difference

```{r db_difference}
referenda %>% 
  mutate(electorate_name = fct_reorder(electorate_name, difference)) %>% 
  plot_ly(color = I("gray80")) %>% 
    add_segments(
      x = ~pmin(cannabis, euthenasia), xend = ~pmax(cannabis, euthenasia),
      y = ~electorate_name, yend = ~electorate_name,
      showlegend = FALSE
    ) %>% 
  add_markers(
    x = ~euthenasia, y = ~electorate_name, name = "euthenasia",
    color = I("gray10"),
    hoverinfo = "text",
    text = ~glue("{round(difference,1)}% {e_compare} voters supported euthenasia in {electorate_name}")
  ) %>% 
  add_markers(
    x = ~cannabis, y = ~electorate_name, name = "cannabis",
    color = I("green"),
    hoverinfo = "text",
    text = ~glue("{round(difference,1)}% {c_compare} voters supported cannabis in {electorate_name}")
  ) %>% 
  layout(
    title = glue("Referenda Support Differences by Electorate"),
    xaxis = list(title = "Votes (Percent)"),
    yaxis = list(title = "", dtick = 1),
    margin = list(l = 130)
  )
```


## Explanation {.sidebar}

This election contained two referenda, one to enact a law legalising euthenasia
(in limited circumstances) and one to support a draft proposal to legalise cannabis.

The euthenasia referendum passed comfortably with 
`r referenda %>% filter(electorate_name == "New Zealand") %>% pull(euthenasia)`% support
but the cannabis referendum narrowly failed with only
`r referenda %>% filter(electorate_name == "New Zealand") %>% pull(cannabis)`% support.

In each case the support varied by electorate - in the dot plots you can see the
variation by electorate (and NZ) for both individually, 
and in the dumbbell plot you can see the variation in the difference of support between them.

Note that ordinary referenda results were not reported separately for the M훮ori electorates,
only for the special votes. That's a bit frustrating as there were definitely different
voting patterns, but there's not much we can do. I've combined the general electorate specials
with the general/M훮ori vote numbers. For the M훮ori special votes I've combined them into a
single entry - combined they only make
`r round(100 * maori_specials_total / total_votes, 1)`%
of the national vote.

